Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-03-19 introduction.tex
Overleaf comments

\emph{Transmission} refers to the process by which an infectious disease generates new infections.
Estimating transmission refers to inferring the parameters of this process, including its components.
The components of transmission (\eg number of contacts per day and age-stratified rates) inform how interventions can be targeted.
They are also important for forecasting future incidence.

@@

The data sources used for outbreak surveillance are often observational in nature.
Therefore, they suffer from biases that can be hard to understand and quantify.
Here I discuss three types of data which are useful for surveillance.
Incidence-type and prevalence-type data can directly inform estimation of incidence and transmission, however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
These quantities can be estimated from other studies, \eg cohort studies.

@@

However, severe symptoms can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2).
This means that the observations do not provide much information on incidence in those subgroups.

@@

Next, are semi-mechanistic models.
These have components of the transmission process in them but do not fully represent the process.
The most common type of semi-mechanistic model are those based on a \emph{renewal equation}.
The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
\begin{align}
    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
\end{align}
where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the instantaneous probability of an individual infected $\tau$ time units ago infects an individual at time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
The effective reproduction number can be constructed as a function of $\beta(t, \tau)$ and the generation interval.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature further in this thesis; for details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.

Finally, mechanistic models are the most complex, representing the transmission process in varying levels of detail.
\emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
More complex models relax these assumptions, in particular incorporating heterogeneity; the age structure of the population is important for respiratory diseases.
% Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
% Unlike renewal equations, there are not widely-used implementations of compartmental models.
% A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
% The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

@@

The \emph{disease model} links the transmission model to a population-level quantity which we will make observations of.
Examples are the incidence or prevalence of symptomatic individuals and admissions to hospital.

The choice of observation model can, in theory, be independent of the transmission model.
However, pragmatic considerations often mean that the observation model is chosen to be in a similar form to the transmission model.

Typically, the core of an observation model is a \emph{convolution equation} or an approximation of it.
This equation describes the expectation of the quantity at time $t$, $W_t$, conditional on the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
    \label{intro:eq:disease-model}
\end{align}
where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that are ever observed; and $f(i)$ is the probability that an infection at time $t-i$ is observed at time $t$, conditional on it ever being observed.
$f$ is known as a \emph{delay distribution}.

>>>>

\emph{Transmission} refers to the process by which an infectious disease generates new infections.
Estimating transmission refers to inferring the parameters of this process.
The components of transmission (\eg the number of contacts each individual makes per day) allow interventions to be targeted.
They are also important for forecasting future incidence.

@@

The data sources used for outbreak surveillance are often observational in nature.
Therefore, they suffer from biases that can be hard to understand and quantify.
Here I discuss three types of data which are useful for surveillance.
Incidence-type and prevalence-type data can directly inform estimation of incidence and transmission, however, this estimation almost always relies on prior knowledge of epidemiological and/or biological quantities.
For example, the time between infection and symptom onset.
These quantities can be estimated from other studies, \eg cohort studies.

@@

However, severe symptoms can be rare in some population subgroups (\eg young, otherwise healthy individuals infected with SARS-CoV-2).
This means that the observations do not provide much information on incidence in those subgroups.Furthermore, the delay between infection and the severe event occurring means that recent infections cannot be estimated.

@@

Next, are semi-mechanistic models.
These have components of the transmission process in them but do not fully represent the process.
The most common type of semi-mechanistic model are those based on a \emph{renewal equation}.
The renewal equation~\autocite{fraserRenewal} for an infectious disease, in the most general form, is:
\begin{align}
    Z(t) &= \int_{0}^{t} \beta(t,\tau) Z(t-\tau) d \tau
\end{align}
where $Z(t)$ is the mean incidence at time $t$ and $\beta(t, \tau)$ is the mean rate an individual infected $\tau$ time units ago infects individuals at time $t$.
Therefore, the renewal equation relates incidence in the past to incidence in the future.
This partially describes the transmission process, hence the name semi-mechanistic.
The effective reproduction number can be constructed as a function of $\beta(t, \tau)$ and the generation interval.
% A scaled measure of $I(t)$ (\eg hospital admissions or community test results) allows estimating $\beta$.
% The most common implementations are the R packages EpiNow2~\autocite{EpiNow2} and EpiEstim~\autocite{EpiEstim}.
Renewal equations do not feature further in this thesis; for details, including practical considerations when using them as the basis for inference, see \textcite{nashEstimating,gosticPractical}.

Finally, mechanistic models are the most complex, representing the transmission process in varying levels of detail.
\emph{Compartmental} models, the most common type of mechanistic model, split the population into compartments, each representing a different state of the disease.
The \emph{dynamics} of the epidemic, how it changes over time, are represented by the transitions between compartments (for more details see \cref{E-SEIR:sec:mechanistic-models}).
Simple mechanistic models use a homogeneous population, with individuals differing only due to their disease state, and exponential distributions for the time in each disease state.
More complex models relax these assumptions, in particular to incorporate heterogeneity.
The age structure of the population is important for respiratory diseases because individuals meet individuals of their own age more frequently.
% Mathematically, the dynamics are described by ODEs (ordinary differential equations) or SDEs (stochastic differential equations).
% Unlike renewal equations, there are not widely-used implementations of compartmental models.
% A commonly used frameork is the SEIR (susceptible, exposed, infectious, recovered) model.
% The SEIR model, has four components (susceptible, exposed, infectious, recovered), with the population divided into these components; it is described in detail in \cref{E-SEIR:sec:mechanistic-models}.

@@

\emph{Disease models} link transmission models to population-level quantities on which we will make observation.
For example, the incidence or prevalence of symptomatic individuals and admissions to hospital.
They will then be recorded through some system (\eg a testing programme) described by the observation model (see next section).

% The choice of observation model can, in theory, be independent of the transmission model.
% However, pragmatic considerations often mean that the observation model is chosen to be in a similar form to the transmission model.

At the core of many observation models is a \emph{convolution equation} or an approximation of it.
This equation describes the expectation of the quantity at time $t$, $W_t$, conditional on the incidence prior to time $t$.
In discrete time, it is:
\begin{align}
    \E(W_t \mid Z_1, Z_2, \dots) &= \zeta \sum_{i=0}^\infty Z_{t-i} f(i)
    \label{intro:eq:disease-model}
\end{align}
where $Z_t$ is the incidence at time $t$; $\zeta$ is the proportion of infections that are ever observed; and $f(i)$ is the probability that an infection at time $t-i$ is observed at time $t$, conditional on it ever being observed.
$f$ is known as a \emph{delay distribution}.

### 2024-03-19 transmission.tex
Overleaf comments

First, in \cref{backcalc}, I use a back-calculation approach, based on \cref{E-inc-prev:eq:EPt}.
This approach enables estimation of incidence within (sub)populations of England.
There are several attractive features about this approach in the context here.
The backcalculation approach makes no assumptions about the transmission process, which have the potential to introduce bias.
It can also easily incorporate arbitrary covariates to adjust for non-response bias.
The scale of the CIS and desire to correct for non-response bias presents challenges for this methodology, especially within the limited computation of the SRS environment.

@@

\subsection{Approach} \label{backcalc:sec:approach}

@@

\subsubsection{Possible approaches}

The first approach, which I call the \emph{incidence-inferred} approach, is a single statistical model where incidence is inferred, with prevalence as a derived quantity.
The basis of the statistical model, in a Bayesian framework, would be as follows.
\begin{align}
    \vec{Z} \mid \vec{\phi} &\dist g_Z(\vec{\phi}) \\
    P_{i,t} &= \matr{S} \vec{Z}_i \\
    Y_{i,t} \mid \vec{P} &\dist \BB(n_{i,t}, P_{i,t})
\end{align}
where $\vec{Z}$ is a vector of incidence proportion for each strata and day defined by some model $g$ with parameters $\vec{\phi}$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\matr{S}$ is the survival matrix defined in \cref{E-inc-prev:eq:Smatrix}; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
The model $g_Z$ would be specified to share information across time and/or strata.
This both helps stabilise the estimation (see\todo{ref wherever I end up explaining backcalculation stability issues}) and allows borrowing of information across strata.
Borrowing information is required because some strata and time combinations have few positives (fewer than five per day).
The relationship between $P$ and $Z$ is from \cref{E-inc-prev:eq:EPt-matrix}, although dividing both sides by $\Npop$ to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.

The second approach, which I call the \emph{prevalence-inferred} approach takes a two-step approach.
First, it uses the statistical model from \cref{E-biology-data:sec:MRP} to estimate prevalence.
Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{E-inc-prev:eq:forward-substitute}.
At a high-level, the statistical model is:
\begin{align}
    \vec{P} \sim \vec{\phi} &\dist g_P(\vec{\phi}) \\
    Y_{i,t} &\dist \BB(n_{i,t}, P_{i,t})
\end{align}
where $g_P$ is analogous to $g_Z$ in providing a joint prior for the prevalence in each strata and day combination.

\subsubsection{Desirable properties}

The incidence-inferred approach is more parsimonious in being a single step procedure.
However, it poses several pragmatic challenges which means I favour the prevalence-inferred approach.
Here I explain that further developments, beyond the scope of this thesis, would be required to use the incidence-inferred approach with a suitable $\pi_Z$ within the computational constraints of the SRS.

@@

The chosen approach should have the following properties.
\begin{enumerate}
    \item The smoothing model over time should penalise changes from continuous exponential growth.
        Penalising changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
        Furthermore, such penalisation has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}.
    \item Limit estimates to epidemiologically and biologically plausible values.
        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
        % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias.
    \item Borrow information across the different covariates.
        The borrowing should penalise changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
        A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) over short periods of time with negligible behavioural and immunity changes.
    \item Be computationally efficient
        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
\end{enumerate}

Properties 1 and 2 suggest a model of the form $\logit(\pi_{i,t}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalised (\ie changes away from linear are penalised).
The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.
Combined with property 4, the form of $g$ the null model is $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified; that is, changes away from this model should be penalised.

@@

For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value (see \cref{SEIR:fig:true-vs-posterior}).
This suggests that these parameters are not practically identifiable in this setting.
The posteriors align very closely with the priors (not shown) which reinforces this finding.
The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics{SEIR/sim/true_vs_posterior}}
    \caption[True vs posterior parameter values]{%
        True vs posterior median parameter values for each parameter in each simulation.
        The black line shows the line $y = x$ which the dots would lie on if the parameter is fully recovered.
        The blue line is a linear regression line.
        For $i^+$, $q_c$, and $\psi$, the posterior recovers the true values.
        For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value.
    }
    \label{SEIR:fig:true-vs-posterior}
    \todo[inline]{Fix x-axis labels overlapping / being unreadable}
\end{figure}

>>>>

First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
This approach enables estimation of incidence within (sub)populations of England.
There are several attractive features about this approach in the context here.
The backcalculation approach makes no assumptions about how transmission occurs, which have the potential to introduce bias.
It can also easily incorporate arbitrary covariates to adjust for non-response bias.
The scale of the CIS and desire to correct for non-response bias presents challenges for this methodology, especially within the limited computation of the SRS environment.

@@

\subsection{Approaches} \label{backcalc:sec:approach}

@@

% \subsubsection{Possible approaches}

The first approach, which I call the \emph{incidence-inferred} approach, is a single statistical model where incidence is inferred, with prevalence as a derived quantity.
The basis of the statistical model, in a Bayesian framework, would be as follows.
\begin{align}
    \vec{Z} \mid \vec{\phi} &\dist g_Z(\vec{\phi}) \\
    \vec{P_{i}} &= \matr{S} \vec{Z}_i \\
    Y_{i,t} \mid \vec{P} &\dist \BB(n_{i,t}, P_{i,t})
\end{align}
where $\vec{Z}$ is a vector of incidence proportion for each strata and day defined by some model $g$ with parameters $\vec{\phi}$; $P_{i,t}$ is the prevalence in stratum $i$ on day $t$; $\matr{S}$ is the survival matrix defined in \cref{E-inc-prev:eq:Smatrix}; $Y_{i,t}$ is the number of positive tests in stratum $i$ on day $t$; and $n_{i,t}$ is the number of tests in stratum $i$ on day $t$.
The model $g_Z$ would be specified to share information across time and/or strata.
This both helps stabilise the estimation (see\todo{ref wherever I end up explaining backcalculation stability issues}) and allows borrowing of information across strata.
Borrowing information is required because some strata and time combinations have few positives (fewer than five per day).
The relationship between $P$ and $Z$ is from \cref{E-inc-prev:eq:EPt-matrix}, although dividing both sides by $\Npop$ to have the values in terms of incidence proportion and prevalence rather than incidence and number of prevalent individuals.

The second approach, which I call the \emph{prevalence-inferred} approach takes a two-step approach.
First, it uses the statistical model from \cref{E-biology-data:sec:MRP} to estimate prevalence.
Then, it transforms the posterior distribution of prevalence to a posterior distribution of incidence using the forward substitution as in \cref{E-inc-prev:eq:forward-substitute}.
At a high-level, the statistical model is:
\begin{align}
    \vec{P} \mid \vec{\phi} &\dist g_P(\vec{\phi}) \\
    Y_{i,t} \mid \vec{P} &\dist \BB(n_{i,t}, P_{i,t})
\end{align}
where $g_P$ is analogous to $g_Z$ in providing a joint prior for the prevalence in each strata and day combination.
Step 2 is...

\subsection{Desirable properties}

The incidence-inferred approach is more parsimonious in being a single step procedure.
However, it poses several pragmatic challenges which means I favour the prevalence-inferred approach.
Here I explain that further developments, beyond the scope of this thesis, would be required to use the incidence-inferred approach with a suitable $g_Z$ within the computational constraints of the SRS.

@@

The chosen approach should have the following properties.
\begin{enumerate}
    \item The smoothing model, $g_Z$ or $g_P$, over time should penalise changes from continuous exponential growth of $\pi$ over time.
        Penalising changes from continuous exponential growth is a natural formulation because mechanistic models predict exponential growth over short periods of time.
        Furthermore, such penalisation has previously been shown to improve estimates of SARS-CoV-2 prevalence by balancing over and under-fitting~\autocite{ealesAppropriately}. Used by CIS (section 3.?).
    \item Limit estimates to epidemiologically and biologically plausible values.
        That is, incidence proportion and prevalence should be in the interval $[0, 1]$.
        % While incidence proportion can theoretically be greater than 1, this would imply that some individuals were infected multiple times in a day which is not biologically plausible.
    \item Have sufficient parameters to incorporate all necessary covariates to correct for non-response bias.
    \item Borrow information across the different covariates.
        The borrowing should penalise changes in the distribution of incidence/prevalence across strata in the same geographic region being constant.
        A constant distribution is predicted by the theoretical study of epidemics, for example the dynamics of stratified, mechanistic transmission models (such as the one introduced in \cref{E-SEIR:sec:structured-populations}) over short periods of time with negligible behavioural and immunity changes.
    \item Be computationally efficient.
        The model will be fit to granular data and therefore needs to be run within the SRS, with its computational limitations (see \cref{E-biology-data:sec:SRS}).
\end{enumerate}

Properties 1 and 2 suggest a model of the form $\logit(\pi_{i,t}) = g(i, t)$, where $g$ is a function with changes in the first derivative penalised w.r.t.\ $t$ (\ie changes away from linear are penalised).
The $\logit$ link function both limits $\pi_{i,t}$ to the interval $[0, 1]$, and logit-linear growth approximates exponential growth for small $\pi_{i,t}$~\autocite{ealesAppropriately}.
Combined with property 4, the form of $g$ the null model is $g(i, t) = a_0 + a_t t + f(i)$ where $a_0$ and $a_t$ are model parameters and $f$ is a function to be specified; that is, changes away from this model should be penalised.

@@

For $\vec{\pi}$ and $\rho$, the correlation between the posterior median and the generative value is weak (see \cref{SEIR:fig:true-vs-posterior}).
This suggests that these parameters are not practically identifiable in this setting.
The posteriors align very closely with the priors (not shown) which reinforces this finding.
The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics{SEIR/sim/true_vs_posterior}}
    \caption[True vs posterior parameter values]{%
        True vs posterior median parameter values for each parameter in each simulation.
        The black line shows the line $y = x$ which the dots would lie on if the parameter is fully recovered.
        The blue line is a linear regression line.
        For $i^+$, $q_c$, and $\psi$, the posterior recovers the true values.
        For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value.
    }
    \label{SEIR:fig:true-vs-posterior}
    \todo[inline]{Fix x-axis labels overlapping / being unreadable}
\end{figure}
